Bucket-Sorted Independent Sets for Algebraic Multigrid

### 2007-07-31 bsis_paper/paper.tex
started paper.tex with place holders for the images in the images directory
git-svn-id: file:///home/rblake/svn/pamg/trunk@9 c7f3fe13-8c30-0410-9127-9a9c42fe009c

\chapter{Bucket Sorted Independent Sets}
\label{chap:BSIS}

@@

\section{CLJP-c}
Recall that CLJP-c colors the graph of $S$ before selecting
$C$-points, and the colors are used as one component of the vertex
weights. As a result, the structure of the coarse grids selected is
improved. More formally, the use of graph coloring provides the
following important result.
\begin{theorem}
\label{5:theorem:CLJPc-neighbor-independence}
For all pairs of vertices $i$ and $j \in \mathcal{N}_i$ CLJP-c
guarantees $w_i \ne w_j$.
\end{theorem}
\begin{proof}
Assume two adjacent vertices $i$ and $j$ have the same weight. That is,
$|S_i^T| = |S_j^T|$ and the weight augmentation provided through
coloring is the same for $i$ and $j$. The graph of $S$, however, is
colored such that $i$ and $j$ are different colors for all $j \in
\mathcal{N}_i$, so a contradiction is reached.
\end{proof}
Theorem~\ref{5:theorem:CLJPc-neighbor-independence} establishes that
all adjacent vertices have different weights in CLJP-c, which is not
guaranteed in CLJP, although is unlikely to occur. The following
corollaries are a result of
Theorem~\ref{5:theorem:CLJPc-neighbor-independence}.
\begin{corollary}
Any set of vertices in the graph of $S$ that share the same weight is
an independent set.
\end{corollary}
\begin{corollary}
\label{5:cor:max-weight-indep-set}
The set of vertices with the largest weight in the graph of $S$ form
an independent set satisfying~\ref{5:eqn:D-conditions}. That is, each
vertex in that independent set has a uniquely maximal weight in its
neighborhood.
\end{corollary}
The first corollary states that independent sets can be selected in
the graph simply by selecting sets of vertices with the same
weight. Corollary~\ref{5:cor:max-weight-indep-set} refines this
observation to a subset of vertices guaranteed to satisfy the
selection criterion. In particular, it shows it is possible to build
the coarse grid by selecting vertices with the maximum weight,
updating weights, selecting the next set of vertices with maximum
weight, and so on. This approach is taken by the algorithm developed
in Section~\ref{5:sec:contribution}. It is proven in
Section~\ref{5:sec:cg-invariance} that despite the difference in
approach, the resulting coarse grid is the same as that selected by
CLJP-c.

@@

The search step in CLJP is implemented as shown in
Algorithm~\ref{5:alg:search}.
\begin{algorithm}[h]
\caption{Coarse-Grid Selection Graph Search}
\label{5:alg:search}
\textsc{Search-Graph}$(S,\, C,\, F)$ \{
\begin{algorithmic}[1]
\STATE $D \leftarrow \emptyset$
\FORALL{$i \notin (C \cup F)$}
\IF{$w_i > w_j,\,\forall j \in \mathcal{N}_i$}
\STATE $D \leftarrow D \cup \{i\}$
\ENDIF
\ENDFOR
\RETURN $D$
\end{algorithmic}
\}
\end{algorithm}
In the first iteration, Line~3 is run $2|E|$ times. The total cost of
search in constructing the coarse grid depends on the number of
iterations needed. Even in the best case of $\Omega(E)$ time, the cost
is significant when the graph contains large numbers of edges, as
usually happens on the lower levels in the grid hierarchy
(see~\cite{alber-PCGS} for examples). In the next section, a new
technique is introduced for conducting the search in coarse-grid
selection algorithms independent of the number of edges in the graph.

Pseudo-code for the weight update in CLJP is shown in Algorithm~\ref{5:alg:setup}.
\begin{algorithm}[h]
\caption{CLJP Weight Update for CSR Matrix}
\label{5:alg:setup}
\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$ \{
\begin{algorithmic}[1]
\FORALL{$d \in D$}
\FORALL{$i \in S_d$}
\STATE $w_i \leftarrow w_i - 1$ \COMMENT{see
  Figure~\ref{3:fig:CLJP-update-heuristics} left}
\STATE $S_d = S_d \setminus \{i\}$ \COMMENT{removing edge from graph}
\ENDFOR
\ENDFOR
\FORALL{$i \notin (C \cup F)$}
\FORALL{$k$ originally in $S_i$}
\IF{$k \in D$}
\STATE mark $k$ NEW-C-POINT
\ENDIF
\ENDFOR
\FORALL{$j \in S_i$}
\IF{$j \notin D$}
\FORALL{$k \in S_j$}
\IF[$i$ and $j$ mutually influenced by $k$]{$k$ is marked NEW-C-POINT}
\STATE $w_j \leftarrow w_j - 1$ \COMMENT{see
  Figure~\ref{3:fig:CLJP-update-heuristics} right}
\STATE $S_i = S_i \setminus \{j\}$ \COMMENT{remove edge from $j$ to $i$}
\ENDIF
\ENDFOR
\ENDIF
\ENDFOR
\FORALL{$k$ originally in $S_i$}
\IF{$k \in D$}
\STATE unmark $k$
\STATE $S_i = S_i \setminus \{k\}$ \COMMENT{remove edge from $k$ to $i$,
  if present}
\ENDIF
\ENDFOR
\ENDFOR
\end{algorithmic}
\}
\end{algorithm}
The level of complication in this update routine is due to the CSR
format and the need to find vertices strongly influenced by new
$C$-points. When a new $C$-point $k$ is selected, the first type of
weight update is trivial since determining the vertices in $S_k$ is
inexpensive. The second type of update is more expensive since the
vertices influenced by $k$ are difficult to determine in a CSR
format. The update requires a search of many vertices $i$ and all of
their strong influencing neighbors $j$. The routine then searches
strongly influencing $j$ to determine if any $k \in D$ strongly
influences both $i$ and $j$. The cost increases dramatically as the
density of $S$ increases. Large operator complexities have a
disproportionately large impact on coarse-grid selection run time. In
the next section, a modified update routine to compliment the new
search technique is introduced.

@@

\subsection{Coarse Grid Invariance}
\label{5:sec:cg-invariance}
Theory is developed in this section to prove coarse-grid selection
invariance in general independent set-based algorithms. The algorithms
considered thus far select independent sets using
(\ref{5:eqn:D-conditions}), meaning $i$ is eligible to be in $D$ if
its weight is larger than the weights of vertices in its neighborhood
$\mathcal{N}_i$. Recall the neighborhood of $i$
(Definition~\ref{3:def:neighborhood}) is the set of vertices strongly
influenced by $i$ or strongly influencing $i$. Algorithms relying on
different and larger neighborhoods, such as \emph{distance-$d$
neighborhoods}, are conceivable.
\begin{definition}
The \emph{distance-$d$ neighborhood} of $i$, denoted
$\mathcal{N}_i^d$, is the set of vertices within $d$ hops of $i$ in
the symmetrized strength matrix, excluding $i$. That is,
$\mathcal{N}_i^d = \left[\bigcup_{j \in
\mathcal{N}_i^{d-1}}\left(\mathcal{N}_j \cup \{j\}\right)\cup
\mathcal{N}_i\right] \setminus \{i\}$, where $d > 0$ and
$\mathcal{N}_i^0 = \emptyset$.
\end{definition}

@@

Given a symmetric, but otherwise arbitrary, set of selection
neighborhoods, the set of vertices potentially able to affect the
vertex weight of $i$ or $j \in \mathcal{N}_i^s$ is the \emph{extended
selection neighborhood}.
\begin{definition}
The \emph{extended selection neighborhood} of $i$, denoted
$\mathcal{N}_i^{2s}$, is the union of the selection neighborhood of
$i$ with the selection neighborhoods of the vertices in
$\mathcal{N}_i^s$, excluding $i$. That is, $\mathcal{N}_i^{2s} =
\left[\left(\bigcup_{j \in \mathcal{N}_i^s}\mathcal{N}_j^s\right)\cup
\mathcal{N}_i^s\right] \setminus \{i\}$.
\end{definition}

When a vertex $i$ satisfies the generalized selection
condition~(\ref{5:eqn:general-selection}), no other $C$-point
assignments affect $w_i$. This is formalized below.
\begin{lemma}
\label{5:lemma:C-lock}
If (\ref{5:eqn:general-selection}) is satisfied for vertex $i$, then
$i$ must be the next vertex in $\{i\} \cup \mathcal{N}_i^s$ to become
a $C$-point, regardless of any other selections and updates made in
the graph.
\end{lemma}
\begin{proof}
The satisfaction of (\ref{5:eqn:general-selection}) means $w_i > w_j$
for all $j \in \mathcal{N}_i^s$. For $i$ to not become a $C$-point,
its weight must become smaller than some $j \in \mathcal{N}_i^s$. The
weight of $i$, however, is not decremented unless some $j \in
\mathcal{N}_i^s$ satisfies~(\ref{5:eqn:general-selection}) and becomes
a $C$-point, which is impossible until after $i$ is assigned to the
$C$-point set.
\end{proof}

To demonstrate all algorithms using the same selection neighborhood
and update rules select identical coarse grids, a proof with an
inductive argument is presented below. The base case is provided by
the first set of vertices satisfying the general selection conditions.
\begin{definition}
Let $D_0$ be the set of all vertices satisfying
(\ref{5:eqn:general-selection}) in the first iteration of coarse-grid
selection.
\end{definition}

Vertices in $D_0$ are destined to become $C$-points regardless of the
algorithm used to build coarse grids. In the independent set-based
algorithms discussed in Chapters~\ref{chap:CGS}
and~\ref{chap:color-based}, all $D_0$ vertices become $C$-points in
the first iteration. Any algorithm constructed, however, eventually
selects all $D_0$ vertices as $C$-points, as proven in the following
lemma.
\begin{lemma}
\label{5:lemma:D0_invariant}
Given the same selection neighborhoods and update heuristics, all
algorithms select vertices in $D_0$ as $C$-points.
\end{lemma}
\begin{proof}
The proof follows from Lemma~\ref{5:lemma:C-lock}. All vertices in
$D_0$ satisfy the selection condition~(\ref{5:eqn:general-selection}),
so the assignment of any other $C$-point has no effect on the weight
of a $D_0$ vertex.
\end{proof}

@@

\begin{theorem}
\label{5:theorem:invariance}
All independent set-based coarse-grid selection algorithms given the
same initial weights and using the same selection neighborhood,
selection criterion based on~(\ref{5:eqn:general-selection}), and
weight update heuristics as described above select identical coarse
grids.
\end{theorem}
\begin{proof}
Let $c$ be the vertices in $\mathcal{N}_i^{2s}$ satisfying the
conditions for $D$ in some arbitrary iteration. Suppose assigning
vertices $c_1 \subset c$ to the $C$-point set leads to $w_i > w_j$ for
all $j \in \mathcal{N}_i^s$. Also suppose assigning $c_2 \subset c$,
$c_1 \ne c_2$, to the $C$-point set leads to the existence of some $j
\in \mathcal{N}_i^s$ such that $w_j > w_k$ for all $k \in
\mathcal{N}_j^s$.

@@

\begin{remark}
CLJP and CLJP-c use the distance-one neighborhood as the selection
neighborhood, (\ref{5:eqn:D-conditions}) as the selection criterion,
and the weight update heuristics in
Algorithm~\ref{3:alg:CLJP-update-heuristics}. Given the same initial
weights, all algorithms based on the parameters utilized by CLJP and
CLJP-c select identical coarse grids.
\end{remark}

@@

In each iteration, the non-empty bucket with largest weight forms $D$
(see Corollary~\ref{5:cor:max-weight-indep-set}). These vertices are
assigned to the $C$-point set and removed from the data
structure. Vertex weight updates lead to corresponding updates to the
data structure, and new $F$-points are removed from the data
structure. These operations continue until all buckets are empty, at
which point the coarse-grid selection is
complete. Algorithms~\ref{5:alg:BSIS-setup},
\ref{5:alg:BSIS-indep-set-select}, and~\ref{5:alg:BSIS-weight-update}
outline the operations discussed above.
\begin{algorithm}[!h]
\caption{BSIS Data Structure Setup}
\label{5:alg:BSIS-setup}
\textsc{BSIS-Setup}$(S)$ \{
\begin{algorithmic}[1]
\FORALL{$i \in V$}
\STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
\STATE $bucket[bucketID].\textsc{Insert}(i)$
\ENDFOR
\end{algorithmic}
\}
\end{algorithm}
\begin{algorithm}[!h]
\caption{Independent Set Selection}
\label{5:alg:BSIS-indep-set-select}
\textsc{BSIS-Independent-Set-Selection}$(S)$ \{
\begin{algorithmic}[1]
\RETURN non-empty $bucket$ with largest $bucketID$
\end{algorithmic}
\}
\end{algorithm}
\begin{algorithm}[!h]
\caption{BSIS Weight Update}
\label{5:alg:BSIS-weight-update}
\textsc{BSIS-Weight-Update}$(S)$ \{
\begin{algorithmic}[1]
\STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
\STATE $bucket[bucketID].\textsc{Remove}(i)$
\STATE $bucket[bucketID - numColors].\textsc{Insert}(i)$
\end{algorithmic}
\}
\end{algorithm}
\begin{figure}
  \begin{center}
    %\includegraphics[width=0.8\textwidth]{images/5/bucket-algorithm.eps}
    \includegraphics[width=0.98\textwidth]{images/5/BSIS/initial.eps}
    \caption{The BSIS data structure.}
    \label{5:fig:bucket-algorithm}
  \end{center}
\end{figure}

Figure~\ref{5:fig:bucket-algorithm-cp1} illustrates the graph and data
structure following the first iteration of the algorithm. Vertex~10
has become a $C$-point and its neighbors weights have been
updated. Vertices assigned to $F$ or $C$ are removed from the data
structure and other affected vertices are moved to new locations in
the data structure. Such vertices are highlighted in red.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/5/BSIS/cp1.eps}
    \caption{The BSIS data structure after selecting the first
    $C$-point (Vertex 10). The weights of neighbors of the new
    $C$-point are updated. Some neighbors become $F$-points and are
    removed from the data structure. Vertices removed from the data
    structure are highlighted with a red ring in the graph, while
    other neighbors are moved to new locations in the data structure
    and are highlighted (to help in reading the figure) in the data
    structure with a red box.}
    \label{5:fig:bucket-algorithm-cp1}
  \end{center}
\end{figure}

@@

Figure~\ref{5:fig:cp1-agg} depicts the data structure after the first
set of $C$-points is selected.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/5/BSIS/cp1-agg.eps}
    \caption{The BSIS data structure after selecting the first
    $C$-point (Vertex 10) with aggregate weight updates.}
    \label{5:fig:cp1-agg}
  \end{center}
\end{figure}
Rather than moving vertices to new buckets, the method now keeps them
in the same location and only moves them when necessary. As shown in
Section~\ref{5:sec:experiments}, aggregation of the weight updates
leads to significant savings in computational cost.

@@

>>>>

\documentclass{elsart}

% Use the option doublespacing or reviewcopy to obtain double line spacing
% \documentclass[doublespacing]{elsart}

% if you use PostScript figures in your article
% use the graphics package for simple commands
% \usepackage{graphics}
% or use the graphicx package for more complicated commands
\usepackage[draft]{graphicx}
% or use the epsfig package if you prefer to use the old commands
% \usepackage{epsfig}

\usepackage{amssymb,listings}

%% THE ENVIRONMENT \begin{proof} ... \end{proof} PRODUCES AN
%% END-OF-PROOF SIGN.
\newenvironment{proof}{\begin{pf}}{\qed\end{pf}}
%% THE ENVIRONMENT \begin{proof*} ... \end{proof*} SUPPORTS ADDING
%% INFORMATION SUCH AS  'Proof (Sketch)' AS THE LABEL FOR THE PROOF.
\newenvironment{proof*}{\begin{pf*}}{\qed\end{pf*}}

% The lineno packages adds line numbers. Start line numbering with
% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
% for the whole article with \linenumbers.
% \usepackage{lineno}
% \linenumbers

\usepackage{listings}
\lstset{mathescape,
        escapechar=\%,
        numbers=left, 
        numberstyle=\tiny, 
        stepnumber=1, 
        numbersep=5pt,
        frame=tb,
        xleftmargin=5mm,
        framexleftmargin=5mm,
        language=Matlab,
        belowcaptionskip=5pt}
\renewcommand{\lstlistingname}{Algorithm}
\usepackage{caption}
\captionsetup[lstlisting]{justification=raggedright}

\begin{document}
\begin{frontmatter}

% Title, authors and addresses

% use the thanksref command within \title, \author or \address for footnotes;
% use the corauthref command within \author for corresponding author footnotes;
% use the ead command for the email address,
% and the form \ead[url] for the home page:
\title{Bucket-Sorted Independent Sets for Algebraic Multigrid}
\author{David Alber\corauthref{cor}},
\corauth[cor]{Corresponding author. Address: 201 N. Goodwin Ave., Siebel
Center for Computer Science, Urbana, IL 61801}
\ead{dalber@uiuc.edu}
\ead[url]{http://www.cs.uiuc.edu/homes/dalber}
\author{Luke Olson}
\address{Department of Computer Science\\ University of Illinois at Urbana-Champaign}
\ead{dalber@uiuc.edu}
\ead[url]{http://www.cs.uiuc.edu/homes/lukeo}
% \thanks[label2]{}
% \address{Address\thanksref{label3}}
% \thanks[label3]{}
% use optional labels to link authors explicitly to addresses:
% \author[label1,label2]{}
% \address[label2]{}

\begin{abstract}
% Text of abstract
\end{abstract}

\begin{keyword}
multigrid \sep parallel algorithms
% keywords here, in the form: keyword \sep keyword

% PACS codes here, in the form: \PACS code \sep code
% \PACS 
\MSC 65F10 \sep 65Y05 \sep 05C69
\end{keyword}
\end{frontmatter}

% main text

@@

\section{CLJP-c}
Recall that CLJP-c colors the graph of $S$ before selecting
$C$-points, and the colors are used as one component of the vertex
weights. As a result, the structure of the coarse grids selected is
improved. More formally, the use of graph coloring provides the
following important result.
\begin{thm}
\label{5:theorem:CLJPc-neighbor-independence}
For all pairs of vertices $i$ and $j \in \mathcal{N}_i$ CLJP-c
guarantees $w_i \ne w_j$.
\end{thm}
\begin{proof}
Assume two adjacent vertices $i$ and $j$ have the same weight. That is,
$|S_i^T| = |S_j^T|$ and the weight augmentation provided through
coloring is the same for $i$ and $j$. The graph of $S$, however, is
colored such that $i$ and $j$ are different colors for all $j \in
\mathcal{N}_i$, so a contradiction is reached.
\end{proof}
Theorem~\ref{5:theorem:CLJPc-neighbor-independence} establishes that
all adjacent vertices have different weights in CLJP-c, which is not
guaranteed in CLJP, although is unlikely to occur. The following
corollaries are a result of
Theorem~\ref{5:theorem:CLJPc-neighbor-independence}.
\begin{cor}
Any set of vertices in the graph of $S$ that share the same weight is
an independent set.
\end{cor}
\begin{cor}
\label{5:cor:max-weight-indep-set}
The set of vertices with the largest weight in the graph of $S$ form
an independent set satisfying~\ref{5:eqn:D-conditions}. That is, each
vertex in that independent set has a uniquely maximal weight in its
neighborhood.
\end{cor}
The first corollary states that independent sets can be selected in
the graph simply by selecting sets of vertices with the same
weight. Corollary~\ref{5:cor:max-weight-indep-set} refines this
observation to a subset of vertices guaranteed to satisfy the
selection criterion. In particular, it shows it is possible to build
the coarse grid by selecting vertices with the maximum weight,
updating weights, selecting the next set of vertices with maximum
weight, and so on. This approach is taken by the algorithm developed
in Section~\ref{5:sec:contribution}. It is proven in
Section~\ref{5:sec:cg-invariance} that despite the difference in
approach, the resulting coarse grid is the same as that selected by
CLJP-c.

@@

The search step in CLJP is implemented as shown in
Algorithm~\ref{5:alg:search}.
\begin{lstlisting}[caption={Coarse-Grid Selection Graph Search},label=5:alg:search]
%\textsc{Search-Graph}$(S,\, C,\, F)$%
$D \leftarrow \emptyset$
for $i$ in $(C \cup F)$
  if $w_i > w_j,\,\forall j \in \mathcal{N}_i$
    $D \leftarrow D \cup \{i\}$
  end
end
return $D$
\end{lstlisting}
% \begin{algorithm}[h]
% \caption{Coarse-Grid Selection Graph Search}
% \label{5:alg:search}
% \textsc{Search-Graph}$(S,\, C,\, F)$ \{
% \begin{algorithmic}[1]
% \STATE $D \leftarrow \emptyset$
% \FORALL{$i \notin (C \cup F)$}
% \IF{$w_i > w_j,\,\forall j \in \mathcal{N}_i$}
% \STATE $D \leftarrow D \cup \{i\}$
% \ENDIF
% \ENDFOR
% \RETURN $D$
% \end{algorithmic}
% \}
% \end{algorithm}
In the first iteration, Line~3 is run $2|E|$ times. The total cost of
search in constructing the coarse grid depends on the number of
iterations needed. Even in the best case of $\Omega(E)$ time, the cost
is significant when the graph contains large numbers of edges, as
usually happens on the lower levels in the grid hierarchy
(see~\cite{alber-PCGS} for examples). In the next section, a new
technique is introduced for conducting the search in coarse-grid
selection algorithms independent of the number of edges in the graph.

Pseudo-code for the weight update in CLJP is shown in Algorithm~\ref{5:alg:setup}.
\begin{lstlisting}[caption={CLJP Weight Update for CSR Matrix},label=5:alg:setup]
%\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$%
for $d \in D$
  for $i \in S_d$
    $w_i \leftarrow w_i - 1$    //see %Figure~\ref{3:fig:CLJP-update-heuristics} left%
    $S_d = S_d \setminus \{i\}$ //removing edge from graph
  end
end
for $i \notin (C \cup F)$
  for $k$ originally in $S_i$
    if $k \in D$
      mark $k$ NEW-C-POINT
    end
  end
  for $j \in S_i$
    if $j \notin D$
      for $k \in S_j$
        if $k$ is marked NEW-C-POINT      // $i$ and $j$ mutually influenced by $k$
          $w_j \leftarrow w_j - 1$        // see %Figure~\ref{3:fig:CLJP-update-heuristics} right%
          $S_i = S_i \setminus \{j\}$     // remove edge from $j$ to $i$
        end
      end
    end
  end
  for $k$ originally in $S_i$
    if $k \in D$
      unmark $k$
      $S_i = S_i \setminus \{k\}$         // remove edge from $k$ to $i$, if present
    end
  end
end
\end{lstlisting}
%\begin{algorithm}[h]
%\caption{CLJP Weight Update for CSR Matrix}
%\label{5:alg:setup}
%\textsc{Update-Weights}$(S,\, D,\, C,\, F,\, w)$ \{
%\begin{algorithmic}[1]
%\FORALL{$d \in D$}
%\FORALL{$i \in S_d$}
%\STATE $w_i \leftarrow w_i - 1$ \COMMENT{see
%  Figure~\ref{3:fig:CLJP-update-heuristics} left}
%\STATE $S_d = S_d \setminus \{i\}$ \COMMENT{removing edge from graph}
%\ENDFOR
%\ENDFOR
%\FORALL{$i \notin (C \cup F)$}
%\FORALL{$k$ originally in $S_i$}
%\IF{$k \in D$}
%\STATE mark $k$ NEW-C-POINT
%\ENDIF
%\ENDFOR
%\FORALL{$j \in S_i$}
%\IF{$j \notin D$}
%\FORALL{$k \in S_j$}
%\IF[$i$ and $j$ mutually influenced by $k$]{$k$ is marked NEW-C-POINT}
%\STATE $w_j \leftarrow w_j - 1$ \COMMENT{see
%  Figure~\ref{3:fig:CLJP-update-heuristics} right}
%\STATE $S_i = S_i \setminus \{j\}$ \COMMENT{remove edge from $j$ to $i$}
%\ENDIF
%\ENDFOR
%\ENDIF
%\ENDFOR
%\FORALL{$k$ originally in $S_i$}
%\IF{$k \in D$}
%\STATE unmark $k$
%\STATE $S_i = S_i \setminus \{k\}$ \COMMENT{remove edge from $k$ to $i$,
%  if present}
%\ENDIF
%\ENDFOR
%\ENDFOR
%\end{algorithmic}
%\}
%\end{algorithm}
The level of complication in this update routine is due to the CSR
format and the need to find vertices strongly influenced by new
$C$-points. When a new $C$-point $k$ is selected, the first type of
weight update is trivial since determining the vertices in $S_k$ is
inexpensive. The second type of update is more expensive since the
vertices influenced by $k$ are difficult to determine in a CSR
format. The update requires a search of many vertices $i$ and all of
their strong influencing neighbors $j$. The routine then searches
strongly influencing $j$ to determine if any $k \in D$ strongly
influences both $i$ and $j$. The cost increases dramatically as the
density of $S$ increases. Large operator complexities have a
disproportionately large impact on coarse-grid selection run time. In
the next section, a modified update routine to compliment the new
search technique is introduced.

@@

\subsection{Coarse Grid Invariance}
\label{5:sec:cg-invariance}
Theory is developed in this section to prove coarse-grid selection
invariance in general independent set-based algorithms. The algorithms
considered thus far select independent sets using
(\ref{5:eqn:D-conditions}), meaning $i$ is eligible to be in $D$ if
its weight is larger than the weights of vertices in its neighborhood
$\mathcal{N}_i$. Recall the neighborhood of $i$
(Definition~\ref{3:def:neighborhood}) is the set of vertices strongly
influenced by $i$ or strongly influencing $i$. Algorithms relying on
different and larger neighborhoods, such as \emph{distance-$d$
neighborhoods}, are conceivable.
\begin{defn}
The \emph{distance-$d$ neighborhood} of $i$, denoted
$\mathcal{N}_i^d$, is the set of vertices within $d$ hops of $i$ in
the symmetrized strength matrix, excluding $i$. That is,
$\mathcal{N}_i^d = \left[\bigcup_{j \in
\mathcal{N}_i^{d-1}}\left(\mathcal{N}_j \cup \{j\}\right)\cup
\mathcal{N}_i\right] \setminus \{i\}$, where $d > 0$ and
$\mathcal{N}_i^0 = \emptyset$.
\end{defn}

@@

Given a symmetric, but otherwise arbitrary, set of selection
neighborhoods, the set of vertices potentially able to affect the
vertex weight of $i$ or $j \in \mathcal{N}_i^s$ is the \emph{extended
selection neighborhood}.
\begin{defn}
The \emph{extended selection neighborhood} of $i$, denoted
$\mathcal{N}_i^{2s}$, is the union of the selection neighborhood of
$i$ with the selection neighborhoods of the vertices in
$\mathcal{N}_i^s$, excluding $i$. That is, $\mathcal{N}_i^{2s} =
\left[\left(\bigcup_{j \in \mathcal{N}_i^s}\mathcal{N}_j^s\right)\cup
\mathcal{N}_i^s\right] \setminus \{i\}$.
\end{defn}

When a vertex $i$ satisfies the generalized selection
condition~(\ref{5:eqn:general-selection}), no other $C$-point
assignments affect $w_i$. This is formalized below.
\begin{lem}
\label{5:lemma:C-lock}
If (\ref{5:eqn:general-selection}) is satisfied for vertex $i$, then
$i$ must be the next vertex in $\{i\} \cup \mathcal{N}_i^s$ to become
a $C$-point, regardless of any other selections and updates made in
the graph.
\end{lem}
\begin{proof}
The satisfaction of (\ref{5:eqn:general-selection}) means $w_i > w_j$
for all $j \in \mathcal{N}_i^s$. For $i$ to not become a $C$-point,
its weight must become smaller than some $j \in \mathcal{N}_i^s$. The
weight of $i$, however, is not decremented unless some $j \in
\mathcal{N}_i^s$ satisfies~(\ref{5:eqn:general-selection}) and becomes
a $C$-point, which is impossible until after $i$ is assigned to the
$C$-point set.
\end{proof}

To demonstrate all algorithms using the same selection neighborhood
and update rules select identical coarse grids, a proof with an
inductive argument is presented below. The base case is provided by
the first set of vertices satisfying the general selection conditions.
\begin{defn}
Let $D_0$ be the set of all vertices satisfying
(\ref{5:eqn:general-selection}) in the first iteration of coarse-grid
selection.
\end{defn}

Vertices in $D_0$ are destined to become $C$-points regardless of the
algorithm used to build coarse grids. In the independent set-based
algorithms discussed in Chapters~\ref{chap:CGS}
and~\ref{chap:color-based}, all $D_0$ vertices become $C$-points in
the first iteration. Any algorithm constructed, however, eventually
selects all $D_0$ vertices as $C$-points, as proven in the following
lemma.
\begin{lem}
\label{5:lemma:D0_invariant}
Given the same selection neighborhoods and update heuristics, all
algorithms select vertices in $D_0$ as $C$-points.
\end{lem}
\begin{proof}
The proof follows from Lemma~\ref{5:lemma:C-lock}. All vertices in
$D_0$ satisfy the selection condition~(\ref{5:eqn:general-selection}),
so the assignment of any other $C$-point has no effect on the weight
of a $D_0$ vertex.
\end{proof}

@@

\begin{thm}
\label{5:theorem:invariance}
All independent set-based coarse-grid selection algorithms given the
same initial weights and using the same selection neighborhood,
selection criterion based on~(\ref{5:eqn:general-selection}), and
weight update heuristics as described above select identical coarse
grids.
\end{thm}
\begin{proof}
Let $c$ be the vertices in $\mathcal{N}_i^{2s}$ satisfying the
conditions for $D$ in some arbitrary iteration. Suppose assigning
vertices $c_1 \subset c$ to the $C$-point set leads to $w_i > w_j$ for
all $j \in \mathcal{N}_i^s$. Also suppose assigning $c_2 \subset c$,
$c_1 \ne c_2$, to the $C$-point set leads to the existence of some $j
\in \mathcal{N}_i^s$ such that $w_j > w_k$ for all $k \in
\mathcal{N}_j^s$.

@@

\begin{rem}
CLJP and CLJP-c use the distance-one neighborhood as the selection
neighborhood, (\ref{5:eqn:D-conditions}) as the selection criterion,
and the weight update heuristics in
Algorithm~\ref{3:alg:CLJP-update-heuristics}. Given the same initial
weights, all algorithms based on the parameters utilized by CLJP and
CLJP-c select identical coarse grids.
\end{rem}

@@

In each iteration, the non-empty bucket with largest weight forms $D$
(see Corollary~\ref{5:cor:max-weight-indep-set}). These vertices are
assigned to the $C$-point set and removed from the data
structure. Vertex weight updates lead to corresponding updates to the
data structure, and new $F$-points are removed from the data
structure. These operations continue until all buckets are empty, at
which point the coarse-grid selection is
complete. Algorithms~\ref{5:alg:BSIS-setup},
\ref{5:alg:BSIS-indep-set-select}, and~\ref{5:alg:BSIS-weight-update}
outline the operations discussed above.
\begin{lstlisting}[caption={BSIS Data Structure Setup},label=5:alg:BSIS-setup]
%\textsc{BSIS-Setup}$(S)$%
for $i \in V$
  $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
  $bucket[bucketID].\textsc{Insert}(i)$
end
\end{lstlisting}
% \begin{algorithm}[!h]
% \caption{BSIS Data Structure Setup}
% \label{5:alg:BSIS-setup}
% \textsc{BSIS-Setup}$(S)$ \{
% \begin{algorithmic}[1]
% \FORALL{$i \in V$}
% \STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
% \STATE $bucket[bucketID].\textsc{Insert}(i)$
% \ENDFOR
% \end{algorithmic}
% \}
% \end{algorithm}
\begin{lstlisting}[caption={Independent Set Selection},label=5:alg:BSIS-indep-set-select]
%\textsc{BSIS-Independent-Set-Selection}$(S)$%
return non-empty $bucket$ with largest $bucketID$
\end{lstlisting}
% \begin{algorithm}[!h]
% \caption{Independent Set Selection}
% \label{5:alg:BSIS-indep-set-select}
% \textsc{BSIS-Independent-Set-Selection}$(S)$ \{
% \begin{algorithmic}[1]
% \RETURN non-empty $bucket$ with largest $bucketID$
% \end{algorithmic}
% \}
% \end{algorithm}
\begin{lstlisting}[caption={BSIS Weight Update},label=5:alg:BSIS-weight-update]
%\textsc{BSIS-Weight-Update}$(S)$%
$bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
$bucket[bucketID].\textsc{Remove}(i)$
$bucket[bucketID - numColors].\textsc{Insert}(i)$
\end{lstlisting}
%\begin{algorithm}[!h]
%\caption{BSIS Weight Update}
%\label{5:alg:BSIS-weight-update}
%\textsc{BSIS-Weight-Update}$(S)$ \{
%\begin{algorithmic}[1]
%\STATE $bucketID \leftarrow (w_i-1)\cdot numColors + color_i$
%\STATE $bucket[bucketID].\textsc{Remove}(i)$
%\STATE $bucket[bucketID - numColors].\textsc{Insert}(i)$
%\end{algorithmic}
%\}
%\end{algorithm}
\begin{figure}
  \begin{center}
    %\includegraphics[width=0.8\textwidth]{images/bucket-algorithm}
    \includegraphics[width=0.98\textwidth]{images/initial}
    \caption{The BSIS data structure.}
    \label{5:fig:bucket-algorithm}
  \end{center}
\end{figure}

Figure~\ref{5:fig:bucket-algorithm-cp1} illustrates the graph and data
structure following the first iteration of the algorithm. Vertex~10
has become a $C$-point and its neighbors weights have been
updated. Vertices assigned to $F$ or $C$ are removed from the data
structure and other affected vertices are moved to new locations in
the data structure. Such vertices are highlighted in red.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/cp1}
    \caption{The BSIS data structure after selecting the first
    $C$-point (Vertex 10). The weights of neighbors of the new
    $C$-point are updated. Some neighbors become $F$-points and are
    removed from the data structure. Vertices removed from the data
    structure are highlighted with a red ring in the graph, while
    other neighbors are moved to new locations in the data structure
    and are highlighted (to help in reading the figure) in the data
    structure with a red box.}
    \label{5:fig:bucket-algorithm-cp1}
  \end{center}
\end{figure}

@@

Figure~\ref{5:fig:cp1-agg} depicts the data structure after the first
set of $C$-points is selected.
\begin{figure}
  \begin{center}
    \includegraphics[width=0.98\textwidth]{images/cp1-agg}
    \caption{The BSIS data structure after selecting the first
    $C$-point (Vertex 10) with aggregate weight updates.}
    \label{5:fig:cp1-agg}
  \end{center}
\end{figure}
Rather than moving vertices to new buckets, the method now keeps them
in the same location and only moves them when necessary. As shown in
Section~\ref{5:sec:experiments}, aggregation of the weight updates
leads to significant savings in computational cost.

@@

\end{document}
